\newcommand{\mytitle}{X-ray Absorption Fine-Structure Spectroscopy}
\newcommand{\myauthor}{Matthew Newville}

\begin{center}
  \Large{Fundamental of X-ray Absorption Fine-Structure Spectroscopy}
\end{center}
\vspace{5mm}

\begin{center} Matthew Newville \end{center}
\begin{center} Center for Advanced Radiation Sources\par
University of Chicago, Chicago, IL\par\end{center}

\vspace{5mm}

The basic physical principles of X-ray Absorption Fine-Structure (XAFS) are
presented.  XAFS is an element-specific spectroscopy in which measurements
are made by tuning the X-ray energy at and above a selected core-level
binding energy of a specified element.  Although XAFS is a well-established
technique providing reliable and useful information about the chemical and
physical environment of the probe atom, its requirement for an
energy-tunable X-ray source means it is primarily done with synchrotron
radiation sources and so is somewhat less common than other spectroscopic
analytical methods.  XAFS spectra are especially sensitive to the oxidation
state and coordination chemistry of the selected element.  In addition, the
extended oscillations of the XAFS spectra are sensitive to the distances,
coordination number and species of the atoms immediately surrounding the
selected element.  This Extended X-ray Absorption Fine-Structure (EXAFS) is
the main focus of this chapter.  As it is element-specific, XAFS places few
restrictions on the form of the sample, and can be used in a variety of
systems and bulk physical environments, including crystals, glasses,
liquids, and heterogeneous mixtures.  Additionally, XAFS can often be done
on low-concentration elements (typically down to a few ppm), and so has
applications in a wide range of scientific fields, including chemistry,
biology, catalysis research, material science, environmental science, and
geology.  Special attention is given here to the basic concepts used in
analysis and modeling of EXAFS spectra.



\thispagestyle{empty}
\clearpage

